Select-based random access \\ to variable-byte encodings

### 2020-04-17 gradu/msc_english_contents.tex
more references

Fundamental method for data compression is variable-length coding \citep{Sal99}. The main idea of variable-length encoding is that 
frequent sequences of data are represented with shorter codewords. Because the sequences of data have different lengths when compressed, it is 
not trivial to determine the exact location of a certain element in the compressed data. If this is required, the usual 
data compression algorithms are inefficient. Fortunately, such random access to elements is not a requirement compression algorithms usually 
need to fulfill. However, random access to compressed data is very useful in compressed data structures. It saves storage space, bandwidth and 
increases the likelihood of data already being in cache \citep{Sch02}. In most compression methods, the only way to do this is to decompress data 
from the beginning.

@@

\section{VB encoding}
VB encoding splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The continuation bit is set to 1 only
on the block containing the least significant bits of the integer. This information is used in decoding to signal whether or not the current integer continues in the next chunk.

For example, the standard bit representation of an unsigned integer 42 is \texttt{00000000 00101010}. When split to blocks of length 4, it becomes \texttt{0000 0000 0010 1010}. The empty blocks 
are removed and then continuation bits are added. 1 is prepended to the least significant block and 0 to the other blocks, resulting in \texttt{00010 11010}. A pseudo code for VB encoding with 
block length 4 is shown in Figure~\ref{vbyte_enc}. \textsc{Prepend} adds an element to the beginning of the list 
and \textsc{extend} adds all the elements of the second list to the end of the first list.

@@

\section{Rank and Select}
Rank and select are two operations on a bit array B. $Rank_1$(B, i) gives the number of 1 bits between B[0] and B[i-1], inclusive, and 
$select_1$(B, i) gives the index of i-th 1 bit in B. Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. To use rank or select in indexing, the
set 1 bits in B should reflect to the element locations in the encoded data.

For most compression algorithms, this requires B to be created in addition to the existing data and the length of B usually 
has to be close to the length of the data, because encoded element lengths vary. VB encoding has several advantages with search and rank: the data is compressed in 
blocks of equal length which significantly shrinks the length of B. Moreover the bit array C formed from the continuation bits already stores the locations of items and works
as the needed indexing array. In this case, $rank_1$(C, i) would give the sum of end blocks up to i-th index and $select_1$(C, i) would give the location of the ending block 
of i-th compressed element.

@@

The rank and select implementations used in this thesis are from C++ library 'SDSL-lite' by \citep{gbmp2014sea}. The library has an implementation of a bit array and several implementations
 of both rank and select to support the bit array. Also a few useful functions from the library were used during the implementation phase. Table~\ref{table:supportsize} has $rank$ and $select$ 
size requirements of implementations used in this experiment. Both rank implementations have a constant space requirement over the bit array, while select's needed size depends on 
the number of 1's in the data. This chapter describes a way to achieve fast random access to variable length codes. The described method also has a low space overhead, increasing the size of the 
compressed sequence by much less than one bit per element. Two versions of $rank$ from the 'SDSL-lite' library are used in this thesis, $rank_v$ and $rank_{v5}$ and the $select$ version is $select_{MCL}$. 
All these were chosen, because they are faste and can be created over a regular bit array.

@@

>>>>

Fundamental method for data compression is variable-length coding \citep{Sal99}. The main idea of variable-length encoding is that 
frequent sequences of data are represented with shorter codewords. Because the sequences of data have different lengths when compressed, it is 
not trivial to determine the exact location of a certain element in the compressed data. If this is required, the usual 
data compression algorithms are inefficient. Fortunately, such random access to elements is not a requirement compression algorithms usually 
need to fulfill.

However, random access to compressed data is very useful in compressed data structures. It saves storage space, bandwidth and 
increases the likelihood of data already being in cache \citep{Sch02}. In most compression methods, the only way to do this is to decompress data 
from the beginning.

@@

\section{VB encoding}
The name \textit{variable-byte} is reflected in how the integers are stored. An array stores all the elements with the same bit length. This is very inefficient if a lot of the values are very small but the 
array also has to support very large integers. VB encoding attempts to eliminate the unneeded leading zeros. The bits in an integer are split into blocks of length $b$ and empty blocks from the beginning are 
discarded. Because the lengths of the integers may now be different, the data cannot be stored as it is. Instead a \textit{continuation bit} is added in front of each block to form a $chunk$. This bit is set 
to 0 on all chunks except for the last one which contains the least significant bits. The continuation bit is used in decoding to signal whether or not the current integer continues in the next chunk.

For example, the standard 16-bit representation of an unsigned integer 42 is \texttt{00000000 00101010}. Assuming the block length is 4, it is split to \texttt{0000 0000 0010 1010}. The empty blocks 
are removed from the beginning and then the continuation bits are added. 1 is added in front of the last block (containing the least significant bits) and 0 to the other blocks, resulting in 
\texttt{\underline{0}0010 \underline{1}1010}. Table~\ref{table:vbytes} contains examples of VB encoded integers with byte length 4.

A pseudo code for VB encoding with block length 7 is shown in Figure~\ref{vbyte_enc}. \textsc{Prepend} adds an element to the beginning of the list and 
\textsc{extend} adds all the elements of the second list to the end of the first list. The block length can be changed by replacing 128 with $2^b$, where $b$ is the desired block length.

@@

\section{Rank and Select}
$Rank$ and $select$ are two bit array operations. $Rank_1$(i) gives the number of bits set to 1 in the bit array before index i. $Select_1$(i) gives the index of i-th bit that is set to 1. 
Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. The two methods are related to each other: when used on a bit array 
$B$ = \texttt{0100 1101 \underline{1}011}, $select$(5) returns 8 (underlined) and therefore $rank$(8) returns 4.

To use rank or select in indexing, the set 1 bits in B should reflect to the element locations in the encoded data. For most compression algorithms, this requires B to be 
created in addition to the existing data and the length of B usually 
has to be close to the length of the data, because encoded element lengths vary. VB encoding has several advantages with search and rank: the data is compressed in 
blocks of equal length which significantly shrinks the length of B. Moreover the bit array C formed from the continuation bits already stores the locations of items and works
as the needed indexing array. In this case, $rank_1$(C, i) would give the sum of end blocks before the i-th index and $select_1$(C, i) would give the location of the ending block 
of i-th compressed element.

@@

The rank and select implementations used in this thesis are from C++ library 'SDSL-lite' by \citep{gbmp2014sea}. The library has an implementation of a bit array and several implementations
 of both rank and select to support the bit array. Also a few useful functions from the library were used during the implementation phase. Table~\ref{table:supportsize} has $rank$ and $select$ 
size requirements of implementations used in this experiment. Both rank implementations have a constant space requirement over the bit array, while select's needed size depends on 
the number of 1's in the data. This chapter describes a way to achieve fast random access to variable length codes. The described method also has a low space overhead, increasing the size of the 
compressed sequence by much less than one bit per element. Two versions of $rank$ from the 'SDSL-lite' library are used in this thesis, $rank_v$ and $rank_{v5}$ and the $select$ version is $select_{MCL}$. 
All these were chosen, because they are fast and can be created over a regular bit array.

@@

 - Chapter 3 start: Direct access to any element.... explain why

 - 3.1 explain more what the bit array is

 - 3.1 faste => fast

 - a real life situation where to use subarray
